Challenges in Developing Great Quasi-Monte Carlo Software

### 2023-02-04 MCQMC2022ChallengesQMCSoftwareArticle.tex
Updates from Overleaf

%%%%Fred start
\usepackage{cite}
\usepackage{xspace}
\usepackage[hyphens]{url}
\providecommand{\HickernellFJ}{Hickernell\xspace}

@@

A QMC approximation of $\mu := \mathbb{E}[f(\boldsymbol{X})]$, $\boldsymbol{X} \sim \mathcal{U}[0,1]^d$ can be implemented in a few steps:
\begin{enumerate}
    \item Draw a sequence of $n$ a low discrepancy (LD) \cite{Nie92, SloJoe94, DicEtal14a, DicEtal22a} nodes,
    $\boldsymbol{x}_1,\dots,\boldsymbol{x}_n$ that mimic $\mathcal{U}[0,1]^d$.
    \item Evaluate the integrand $f$ at these nodes to obtain $f(\boldsymbol{x}_i)$, $i=1,\dots,n$.
    \item Estimate the true mean, $\mu$, by the sample mean
    \begin{equation}
        \hat{\mu} := \frac{1}{n} \sum_{i=1}^n f(\boldsymbol{x}_i).
        \label{eq:mc_average}
    \end{equation}  
\end{enumerate}
However, the practice of QMC is often more complicated.  The original problem may need to be rewritten fit above form and/or to facilitate a good approximation with small $n$.  Practitioners may wish to determine $n$ adaptively to satisfy a desired error tolerance.

@@

\begin{itemize}
\item Correct,

@@

\item Sustainable by a community that owns it,

\item Integrated with related libraries, and

@@

Developing software to implement such enhanced MC strategies poses a number of challenges. First, a development team must determine the value of the software to the users. We discuss why great (Q)MC software in necessary in Section \ref{sec:why_we_need_software}. Second, a team must determine high level objectives for the software; the tenets of great software are described in Section \ref{sec:tenets_of_great_software}. Third, an architecture for the  (Q)MC software should be determined. We propose a modularization of the (Q)MC pipeline in Section \ref{sec:qmc_software_architecture}. Next, a team must choose which programming languages, hardware environments, and distribution platforms they wish to support. We discuss the nuance and challenges in these decisions in Section \ref{sec:language_library_environment}. Finally, a project which hopes to stand the test of time must draw attention and support from the broader (Q)MC community. To this end, we encourage collaborative community development and describe its' best practices in  Section \ref{sec:encouraging_collaborative_developement}.

\section{Why Develop QMC Software} \label{sec:why_we_need_software}

@@

\section{Tenets of Great Software} \label{sec:tenets_of_great_software}

@@

Software should be \emph{accessible} in terms of installation, presentation, navigation, exemplification, and communication. A library should be straightforward to install from a common distribution platform, consistent in user interface, intuitive to navigate to modules and documentation, comprehensible in its' examples, and supportive in user engagement. Software accessibility considerations are further discussed in Section \ref{sec:language_library_environment}.

Libraries should be \emph{sustainable} for future development. The goal should be to have a sufficient user base and developer community for updates and maintenance. Similar to accessibility, sustainability is often overlooked for aspirations discussed later in this section. However, we contend that a sub-optimal but user-friendly library with community engagement may be more impactful than an optimal library which lacks support. We promote collaborative, community driven development further in Section \ref{sec:encouraging_collaborative_developement}.

@@

\section{(Quasi-)Monte Carlo Software Architecture} \label{sec:qmc_software_architecture}

@@

\section{Choosing a Environment} \label{sec:language_library_environment}

@@

\section{Encouraging Collaborative Development} \label{sec:encouraging_collaborative_developement}

@@

\section{Example Collaboration with UMBridge and QMCPy} \label{sec:example_collaboration_umbridge}

@@

\AGSComment{
\tableofcontents
\section*{TODO}
\begin{itemize}
    \item Flesh out Section \ref{sec:encouraging_collaborative_developement}
    \item QMC Software Architecture section: better way then description of components and list of softwares? 
    \item For Section \ref{sec:tenets_of_great_software}, the slides also include 
    \begin{itemize}
        \item  Complete---contain the components or  easily access components in other libraries to solve real, complex problems
        \item Current---include the latest and best algorithms
    \end{itemize}
    which I think fall under the umbrella of software being \emph{integrated} with the community.
    \item update QMCPy citation \cite{QMCPy2020a}
    \item figures?
    \item I have left out or been breif about material from the following slides: QMCPy + FEniCS/Dolfin,  Building a Developer Team for Your Library
\end{itemize}
}

@@

\section{Introduction} \label{sec:intro}
Quasi-Monte Carlo (QMC) methods promise great efficiency gains over independent and identically distributed (IID) Monte Carlo (MC) methods.  In some cases QMC  achieves one hundredth of the error of IID MC in the same amount of time. Often, these efficiency gains are obtained simply by  replacing IID sampling by low discrepancy (LD) sampling, which is the heart of QMC.

@@

\section{Available Software for QMC} \label{sec:available} 
QMC software spans  LD sequence generators, algorithms, and applications.  We review the better known software collections, recognizing that some software overlaps multiple categories.
Software focusing on generating high quality LD sequences  or their generators includes
\begin{description}[format=\textup,format=\textbf]
% https://tex.stackexchange.com/questions/74279/how-to-add-bullets-to-description-lists
	\item[BRODA] Sobol' sequences in C, MATLAB, and Excel \cite{BRODA20a},
	\item[Burkhardt] various QMC software in C++, Fortran, MATLAB, \& Python \cite{Bur20a},
	\item[LatNet Builder] Generating vectors/matrices for lattices and digital nets \cite{LEcEtal22a,LatNet},
	\item[MATLAB] Sobol' and Halton sequences, commercial \cite{MAT9.9},
	\item[MPS] Magic Point Shop, lattices and Sobol' sequences \cite{Nuy17a},
	\item[Owen] Randomized Halton sequences in R \cite{Owe20a},
	\item[PyTorch] Scrambled Sobol' sequences \cite{PyTorch},
	\item[QMC.jl] LD Sequences in Julia \cite{Rob20a}, and
	\item [qrng]  Sobol', Halton, and Korobov sequences in R \cite{QRNG2020}.
\end{description}
Software focusing on QMC algorithms and applications includes
\begin{description}[format=\textup,format=\textbf]
	\item[GAIL] Automatic (Q)MC stopping criteria in MATLAB \cite{ChoEtal20a},
	\item[ML(Q)MC] Multi-Level (Quasi-)Monte Carlo routines in C++, MATLAB, Python, and R \cite{GilesSoft},
	\item[OpenTURNS] Open source initiative for the Treatment of Uncertainties, Risks 'N Statistics in Python \cite{OpenTURNS},
	\item[QMC4PDE] QMC for elliptic PDEs with random diffusion coefficients \cite{KuoNuy16a},
	\item[SSJ] Stochastic Simulation in Java \cite{SSJ}, and
	\item[UQLab] Framework for Uncertainty Quantification in MATLAB \cite{UQLab2014}.
\end{description}

@@

The software libraries referenced in Section \ref{sec:available} provide one or more of these components. QMCPy combines multiple examples of them all under and object oriented framework. Each example is implemented as a concrete class that realizes the properties and methods required by the abstract class for that component. The following sections detail descriptions and use cases for each component and select implementations.

@@

\section{Stopping Criteria} \label{sec:stopping_crit}

@@

QMC, which uses LD sequences, is touted as providing substantially greater computational efficiency compared to IID MC.
Figure \ref{fig:sc_comp} compares the time and sample sizes needed to compute the $5$-dimensional Keister integral \eqref{eq:KeisterIntegral} using IID sequences and LD lattice sequences. Consistent with what is stated in Section \ref{sec:intro}, the error of IID MC is $\calO(n^{-1/2})$, which means that the time and sample size to obtain an absolute error tolerance of $\varepsilon$ is $\calO(\varepsilon^{-2})$.  By contrast, the  error of QMC using LD sequences is $\calO(n^{-1+\epsilon})$, which implies $\calO(\varepsilon^{-1-\epsilon})$ times and sample sizes.  We see that QMC methods often require orders of magnitude fewer samples that MC methods to achieve the same error tolerance.

@@

A helpful features of QMCPy is the ability to print out components instances to see more detailed information. This allows users to quickly see the properties that have been set and defaulted for any given component. This feature was already used to print the data object returned by the stopping criterion's \texttt{integrate} method in Section \ref{sec:stopping_crit}. The following code prints a drifted \texttt{BrownianMotion} instance and displays the connection to an underlying Gaussian measure. 
\lstinputlisting[style=Python]{QMCSoftwareArticle/python/print_bm.txt}

@@

\section{Further Work} \label{sec:further}

>>>>

%%%%Fred start
\include{common-macros.tex}
\usepackage{cite}
\usepackage{xspace}
\usepackage[hyphens]{url}
\providecommand{\HickernellFJ}{Hickernell\xspace}

@@

A QMC approximation of $\mu := \EE[f(\bsX)]$, $\bsX \sim \calU[0,1]^d$ can be implemented in a few steps:
\begin{enumerate}
    \item Draw a sequence of $n$ a low discrepancy (LD) \cite{Nie92, SloJoe94, DicEtal14a, DicEtal22a} nodes,
    $\bsx_1,\dots,\bsx_n$ that mimic $\calU[0,1]^d$.
    \item Evaluate the integrand $f$ at these nodes to obtain $f(\bsx_i)$, $i=1,\dots,n$.
    \item Estimate the true mean, $\mu$, by the sample mean
    \begin{equation}
        \hat{\mu} := \frac{1}{n} \sum_{i=1}^n f(\bsx_i).
        \label{eq:mc_average}
    \end{equation}  
\end{enumerate}
However, the practice of QMC is often more complicated.  The original problem may need to be rewritten fit above form and/or to facilitate a good approximation with small $n$.  Practitioners may wish to determine $n$ adaptively to satisfy a desired error tolerance.

@@

\begin{itemize}
\item Integrated with related libraries (Sect.\ \ref{CDHS_sec:integrated}),

\item Correct,

@@

\item Sustainable by a community that owns it, and

@@

Developing software to implement such enhanced MC strategies poses a number of challenges. First, a development team must determine the value of the software to the users. We discuss why great (Q)MC software in necessary in Section \ref{CDHS_sec:why_we_need_software}. Second, a team must determine high level objectives for the software; the tenets of great software are described in Section \ref{CDHS_sec:tenets_of_great_software}. Third, an architecture for the  (Q)MC software should be determined. We propose a modularization of the (Q)MC pipeline in Section \ref{CDHS_sec:qmc_software_architecture}. Next, a team must choose which programming languages, hardware environments, and distribution platforms they wish to support. We discuss the nuance and challenges in these decisions in Section \ref{CDHS_sec:language_library_environment}. Finally, a project which hopes to stand the test of time must draw attention and support from the broader (Q)MC community. To this end, we encourage collaborative community development and describe its' best practices in  Section \ref{CDHS_sec:encouraging_collaborative_developement}.

\section{Why Develop QMC Software} \label{CDHS_sec:why_we_need_software}

@@

\section{Integrated} \label{CDHS_sec:integrated}
We now discuss characteristics of great QMC software outlined in the introduction.

Expanding on the problem formulation in the introduction, we want to approximate the integral or expectation, $\mu$, well by the sample mean, $\hat{\mu}$:
\begin{gather} 
\label{eq:cubSummary}
	\mu : = \int_\calT g(\bst) \, \lambda(\bst) \, \D \bst  = \bbE[f(\bsX)] = \int_{[0,1]^d} f(\bsx)  \, \D \bsx \approx \frac 1n \sum_{i=0}^{n-1} f(\bsx_i) =: \hat{\mu}_n, \\
 \label{CDHS_eq:error_tol}
 \abs{\mu - \hat{\mu}_n} \le \varepsilon.
\end{gather}
Progressing from the left side of this equation to the right requires several steps.

The \emph{integrand}, $g$, and the non-negative weight, $\lambda$, define the original integral arising from an application.  The function $g$ may represent the discounted payoff of a financial derivative, or the velocity of flow at a point in rock with stochastic porosity.  If $\lambda$ is a probability density function for a random variable $\bsT$, then $\mu = \EE[g(\bsT)]$.

To put $\mu$ into the form that is directly accessible to QMC methods requires a \emph{variable transformation}, $\bsPsi$, such that 
\begin{equation*}
    \Psi((0,1)^d) = \calT, \qquad f(x) = g(\bsPsi(\bsx)) \lambda(\bsPsi(\bsx)) \abs{\frac{\partial \bsPsi}{\partial{\bsx}}}.
\end{equation*}
Such a transformation is typically non-unique, and the choice of a good one is equivalent to importance sampling.

An \emph{LD generator}, such as a lattice, digital net, or Halton point generator supplies the sequence $\bsx_1, \bsx_2, \ldots$.  These sequences may be deterministic or random and are intended to have an empirical distribution that approximates well the uniform distribution.

Finally, a \emph{stopping criterion} is determines the sample size, $n$, required to satisfy the error criterion \eqref{CDHS_eq:error_tol} given in terms of the absolute error tolerance, $\varepsilon$.  One can also imagine error criteria involving a relative error tolerance or a combination of absolute and relative error tolerances.

No one software library contains all of these features, which is why great QMC software must be well integrated into other software libraries.  This may mean large libraries including QMC routines and/or building connections between QMC libraries and other libraries.

The most prevalent generally available QMC software are LD generators. They appear in ACM publications \cite{BraFox88,BraFoxNie92,HonHic00a}, FinDer \cite{PasTra95,FinDer}, BRODA \cite{BRODA20a}, NAG \cite{NAG27}, and MATLAB \cite{MAT9.13}.

LatNetBuilder \cite{LatNet} 
    \item Low discrepancy generators \cite{FriKel02,FriKelweb}, SamplePack \cite{SamplePack}, \cite{GruWeb},  and MatBuilder \cite{paulin2022} 
    \item Sobol' direction numbers \cite{JoeKuo03,JoeKuo08,SobolDirection}
    \item Fast CBC, Magic Point Shop, QMC4PDE and other code since 2004  at \cite{NuyWeb} 
    \item Multi-level software \cite{GilesSoft,GilesQSoft} 
    \item Data-driven error bounds and stopping criteria in GAIL \cite{ChoEtal21a} and QMCPy \cite{QMCPy2020a, ChoEtal22a}  
    \item Uncertainty quantification libraries Dakota \cite{DakotaUsersManual}, UQTk \cite{DebEtal04,UQTk}, and MUQ \cite{MUQ}  have some basic level low discrepancy sampling
    \item QMC framework in Julia since 2019 \cite{QMCJulia} by Robbe and others 
    \item Scrambled Sobol' in SciPy \cite{virtanen2020scipy} and PyTorch \cite{paszke2019pytorch} since several years ago
    \item TensorFlow QMC framework \cite{tfqfQMC2021a} since a year or so ago
    \item CUBA \cite{CUBA}

This requires four components, which we implement as QMCPy classes.

\section{Tenets of Great Software} \label{CDHS_sec:tenets_of_great_software}

@@

Software should be \emph{accessible} in terms of installation, presentation, navigation, exemplification, and communication. A library should be straightforward to install from a common distribution platform, consistent in user interface, intuitive to navigate to modules and documentation, comprehensible in its' examples, and supportive in user engagement. Software accessibility considerations are further discussed in Section \ref{CDHS_sec:language_library_environment}.

Libraries should be \emph{sustainable} for future development. The goal should be to have a sufficient user base and developer community for updates and maintenance. Similar to accessibility, sustainability is often overlooked for aspirations discussed later in this section. However, we contend that a sub-optimal but user-friendly library with community engagement may be more impactful than an optimal library which lacks support. We promote collaborative, community driven development further in Section \ref{CDHS_sec:encouraging_collaborative_developement}.

@@

\section{(Quasi-)Monte Carlo Software Architecture} \label{CDHS_sec:qmc_software_architecture}

@@

\section{Choosing a Environment} \label{CDHS_sec:language_library_environment}

@@

\section{Encouraging Collaborative Development} \label{CDHS_sec:encouraging_collaborative_developement}

@@

\section{Example Collaboration with UMBridge and QMCPy} \label{CDHS_sec:example_collaboration_umbridge}

@@

\AGSComment{
\tableofcontents
\section*{TODO}
\begin{itemize}
    \item Flesh out Section \ref{CDHS_sec:encouraging_collaborative_developement}
    \item QMC Software Architecture section: better way then description of components and list of softwares? 
    \item For Section \ref{CDHS_sec:tenets_of_great_software}, the slides also include 
    \begin{itemize}
        \item  Complete---contain the components or  easily access components in other libraries to solve real, complex problems
        \item Current---include the latest and best algorithms
    \end{itemize}
    which I think fall under the umbrella of software being \emph{integrated} with the community.
    \item update QMCPy citation \cite{QMCPy2020a}
    \item figures?
    \item I have left out or been breif about material from the following slides: QMCPy + FEniCS/Dolfin,  Building a Developer Team for Your Library
\end{itemize}
}

@@

\section{Introduction} \label{CDHS_sec:intro}
Quasi-Monte Carlo (QMC) methods promise great efficiency gains over independent and identically distributed (IID) Monte Carlo (MC) methods.  In some cases QMC  achieves one hundredth of the error of IID MC in the same amount of time. Often, these efficiency gains are obtained simply by  replacing IID sampling by low discrepancy (LD) sampling, which is the heart of QMC.

@@

\section{Available Software for QMC} \label{CDHS_sec:available} 
QMC software spans  LD sequence generators, algorithms, and applications.  We review the better known software collections, recognizing that some software overlaps multiple categories.
Software focusing on generating high quality LD sequences  or their generators includes
\begin{description}[format=\textup,format=\textbf]
% https://tex.stackexchange.com/questions/74279/how-to-add-bullets-to-description-lists
	\item[BRODA] Sobol' sequences in C, MATLAB, and Excel \cite{BRODA20a},
	\item[Burkhardt] various QMC software in C++, Fortran, MATLAB, \& Python \cite{Bur20a},
	\item[LatNet Builder] Generating vectors/matrices for lattices and digital nets \cite{LEcEtal22a,LatNet},
	\item[MATLAB] Sobol' and Halton sequences, commercial \cite{MAT9.9},
	\item[MPS] Magic Point Shop, lattices and Sobol' sequences \cite{Nuy17a},
	\item[Owen] Randomized Halton sequences in R \cite{Owe20a},
	\item[PyTorch] Scrambled Sobol' sequences \cite{PyTorch},
	\item[QMC.jl] LD Sequences in Julia \cite{Rob20a}, and
	\item [qrng]  Sobol', Halton, and Korobov sequences in R \cite{QRNG2020}.
\end{description}
Software focusing on QMC algorithms and applications includes
\begin{description}[format=\textup,format=\textbf]
	\item[GAIL] Automatic (Q)MC stopping criteria in MATLAB \cite{ChoEtal20a},
	\item[ML(Q)MC] Multi-Level (Quasi-)Monte Carlo routines in C++, MATLAB, Python, and R \cite{GilesSoft},
	\item[OpenTURNS] Open source initiative for the Treatment of Uncertainties, Risks 'N Statistics in Python \cite{OpenTURNS},
	\item[QMC4PDE] QMC for elliptic PDEs with random diffusion coefficients \cite{KuoNuy16a},
	\item[SSJ] Stochastic Simulation in Java \cite{SSJ}, and
	\item[UQLab] Framework for Uncertainty Quantification in MATLAB \cite{UQLab2014}.
\end{description}

@@

The software libraries referenced in Section \ref{CDHS_sec:available} provide one or more of these components. QMCPy combines multiple examples of them all under and object oriented framework. Each example is implemented as a concrete class that realizes the properties and methods required by the abstract class for that component. The following sections detail descriptions and use cases for each component and select implementations.

@@

\section{Stopping Criteria} \label{CDHS_sec:stopping_crit}

@@

QMC, which uses LD sequences, is touted as providing substantially greater computational efficiency compared to IID MC.
Figure \ref{fig:sc_comp} compares the time and sample sizes needed to compute the $5$-dimensional Keister integral \eqref{eq:KeisterIntegral} using IID sequences and LD lattice sequences. Consistent with what is stated in Section \ref{CDHS_sec:intro}, the error of IID MC is $\calO(n^{-1/2})$, which means that the time and sample size to obtain an absolute error tolerance of $\varepsilon$ is $\calO(\varepsilon^{-2})$.  By contrast, the  error of QMC using LD sequences is $\calO(n^{-1+\epsilon})$, which implies $\calO(\varepsilon^{-1-\epsilon})$ times and sample sizes.  We see that QMC methods often require orders of magnitude fewer samples that MC methods to achieve the same error tolerance.

@@

A helpful features of QMCPy is the ability to print out components instances to see more detailed information. This allows users to quickly see the properties that have been set and defaulted for any given component. This feature was already used to print the data object returned by the stopping criterion's \texttt{integrate} method in Section \ref{CDHS_sec:stopping_crit}. The following code prints a drifted \texttt{BrownianMotion} instance and displays the connection to an underlying Gaussian measure. 
\lstinputlisting[style=Python]{QMCSoftwareArticle/python/print_bm.txt}

@@

\section{Further Work} \label{CDHS_sec:further}
